---
title: "Data Science HW2"
author: "Sabah Usmani"
date: '`r format(Sys.time(), "%Y-%m-%d")`'
output: github_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, message = FALSE)
```


```{r load_libraries}
library(tidyverse)
library(readxl)
```
### Problem 2
Load Mr Trash Dataset and Clean Data

```{r}

mr_trash_data <- read_excel("data/Trash_Wheel_Collection_Data_updated.xlsx",sheet=1, skip = 1, range = cell_cols("A:N")) %>% 
  janitor::clean_names() %>%
  drop_na(dumpster) %>% #omit columns with no dumpster specific data
  mutate(year = as.numeric(year)) %>%
  mutate(sports_balls = as.integer(sports_balls)) %>% 
  mutate(wheel_type = "mister" ) %>%
  select(dumpster, wheel_type, month:chip_bags,homes_powered, sports_balls)


```
Load Professor Trash Data and Clean Data 

```{r}
prof_trash_data <- read_excel("data/Trash_Wheel_Collection_Data_updated.xlsx",sheet=2, skip = 1, range = cell_cols("A:M")) %>%
  janitor::clean_names() %>%
  drop_na(dumpster) %>% 
  mutate(wheel_type = "prof") %>%
  select(dumpster, wheel_type, month:homes_powered) %>%
  mutate(sports_balls = 0)
  
 
```
Combining the two datasets (mister + prof)

```{r}

trash_data <- bind_rows(mr_trash_data, prof_trash_data)

obs <- trash_data %>%
  nrow()

tot_weight <- trash_data %>%
  pull(weight_tons) %>%
  sum()

prof_tot_weight <- prof_trash_data %>%
  pull(weight_tons) %>%
    sum()

mr_sports_2020 <- mr_trash_data %>%
  filter(year == 2020) %>%
  pull(sports_balls) %>%
  sum()
  
```
There are `r obs` observations in the combined dataset, which includes both mr and professor trash wheel. There is no information about sports balls for Professor trash wheel. Key variables include total weight of trash collected in tonnes (weight_tons), volume of trash collected (cubic yards), and the number of plastic bottles collected (plastic_bottles). The total weight of trash collected by both wheels is `r tot_weight`

The total weight of trash collected by professor trash wheel was `r prof_tot_weight`. 
`r mr_sports_2020` sports balls were collected by Mr Trash wheel in 2020.


### Problem 3 

Load all three datasets 

```{r}

pols_data <- read_csv("data/fivethirtyeight_datasets/pols-month.csv")

unemployment_data <- read_csv("data/fivethirtyeight_datasets/unemployment.csv")

snp_data <- read_csv("data/fivethirtyeight_datasets/snp.csv")

```
Cleaning the data in pols-month

```{r}
pols_data_clean <- pols_data %>%
 separate(mon, into = c("year", "month", "day")) %>%
  mutate(year = as.numeric(year)) %>%
 mutate(month = recode(month, `01` = "january", `02` = "february", `03` = "march", `04` = "april", `05` = "may", `06` = "june", `07` = "july", `08` = "august", `09` = "september", `10` = "october", `11` = "november", `12` = "december")) %>%
  mutate(president = if_else(prez_gop == 1 , "gop", "dem")) %>%
  select(-prez_dem, -prez_gop, -day) %>%
  relocate(year, month, president) 

```
Tidying the data in snp

```{r}
snp_data_clean <- snp_data %>%
 separate(date, into = c("month", "day", "year")) %>% 
  mutate(year = as.numeric(year)) %>% 
  mutate(month = recode(month, `1` = "january", `2` = "february", `3` = "march", `4` = "april", `5` = "may", `6` = "june", `7` = "july", `8` = "august", `9` = "september", `10` = "october", `11` = "november", `12` = "december")) %>%
  select(-day) %>% 
  relocate(year, month) %>%
  mutate(year = if_else(year <= 22, year+2000, year+1900))
  
```
Tidying the unemployment data 

```{r}
unemploy_clean <- unemployment_data %>%
  janitor::clean_names() %>%
    pivot_longer(jan:dec,
               names_to = "month",
               #names_prefix = "bdi_score_",
               values_to = "perc_unemployed", 
               ) %>%
  mutate(month = recode(month, `jan` = "january", `feb` = "february", `mar` = "march", `apr` = "april", `may` = "may", `jun` = "june", `jul` = "july", `aug` = "august", `sep` = "september", `oct` = "october", `nov` = "november", `dec` = "december"))

```
Merging snp into pols 

```{r}

snp_into_pols_into_unemploy <- pols_data_clean %>%
  left_join(snp_data_clean, by = c("year", "month")) %>% 
 left_join(unemploy_clean, by = c("year", "month"))

min_year <- snp_into_pols_into_unemploy %>% pull(year) %>% min()

max_year <- snp_into_pols_into_unemploy %>% pull(year) %>% max()
  

```
The pols_month data includes 822 observations of 9 variables related to the number of national politicians who are democratic or republican at any given time. It includes the number of governors, senators, and, representatives on both the democratic and republican side on associated dates. The snp data contained 787 observations related to closing value of the Standard & Poor’s stock market index (S&P) on specific dates of observation. The unemployment data includes 68 observations and has values percentage unemployment for the month for selected years. 

The resulting joined data set is 822 X 11 tibble. The range of the years is from 1947-2015. The name of key variables include year, month, president (whether it was a gop or democratic president), perct_unemployed. the number of republic or democratic senators is also a helpful figure (sen_gop and sen_dem) to better understand if the president had a house majority at the time. 


### Problem 1

Below we import and clean data from `NYC_Transit_Subway_Entrance_And_Exit_Data.csv`. The process begins with data import, updates variable names, and selects the columns that will be used in later parts fo this problem. We update `entry` from `yes` / `no` to a logical variable. As part of data import, we specify that `Route` columns 8-11 should be character for consistency with 1-7.

```{r}
trans_ent = 
  read_csv(
    "data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
    col_types = cols(Route8 = "c", Route9 = "c", Route10 = "c", Route11 = "c")) %>% 
  janitor::clean_names() %>% 
  select(
    line, station_name, station_latitude, station_longitude, 
    starts_with("route"), entry, exit_only, vending, entrance_type, 
    ada) %>% 
  mutate(entry = ifelse(entry == "YES", TRUE, FALSE))
```

As it stands, these data are not "tidy": route number should be a variable, as should route. That is, to obtain a tidy dataset we would need to convert `route` variables from wide to long format. This will be useful when focusing on specific routes, but may not be necessary when considering questions that focus on station-level variables. 

The following code chunk selects station name and line, and then uses `distinct()` to obtain all unique combinations. As a result, the number of rows in this dataset is the number of unique stations.

```{r}
trans_ent %>% 
  select(station_name, line) %>% 
  distinct
```

The next code chunk is similar, but filters according to ADA compliance as an initial step. This produces a dataframe in which the number of rows is the number of ADA compliant stations. 

```{r}
trans_ent %>% 
  filter(ada == TRUE) %>% 
  select(station_name, line) %>% 
  distinct
```

To compute the proportion of station entrances / exits without vending allow entrance, we first exclude station entrances that do not allow vending. Then, we focus on the `entry` variable -- this logical, so taking the mean will produce the desired proportion (recall that R will coerce logical to numeric in cases like this).

```{r}
trans_ent %>% 
  filter(vending == "NO") %>% 
  pull(entry) %>% 
  mean
```

Lastly, we write a code chunk to identify stations that serve the A train, and to assess how many of these are ADA compliant. As a first step, we tidy the data as alluded to previously; that is, we convert `route` from wide to long format. After this step, we can use tools from previous parts of the question (filtering to focus on the A train, and on ADA compliance; selecting and using `distinct` to obtain dataframes with the required stations in rows).

```{r}
trans_ent %>% 
  pivot_longer(
    route1:route11,
    names_to = "route_num",
    values_to = "route") %>% 
  filter(route == "A") %>% 
  select(station_name, line) %>% 
  distinct

trans_ent %>% 
  pivot_longer(
    route1:route11,
    names_to = "route_num",
    values_to = "route") %>% 
  filter(route == "A", ada == TRUE) %>% 
  select(station_name, line) %>% 
  distinct
```

```



